Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-01-26 analysis/paper/London_HauloutBeahvior.Rmd
PLB COMMENTS ADDRESSED

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering, Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance from large scale surveys of these species are unknown, so it is
difficult to assess the effect, if any, declines in sea ice habitat have had, or
will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Thus,
bio-logging devices are especially useful tools for collecting key information
on movement and haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about life history and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi Sea. These studies were often further limited by logistical constraints
on fieldwork and the attachment duration or operational life of bio-loggers. In
this study, we address some of these limitations by deploying small bio-loggers
designed for longer-term attachment on rear flippers of a subset of the study
individuals. These devices are designed to collect data through the molt period
(when those adhered to the hair would fall off) and, in some situations, provide
multiple years of data.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we established baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers attached during ship-based
capture trips and from coastal Alaska communities. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
to the primary literature for detailed capture and bio-logger attachment methods
(Supplemental Material, Table).

@@

Haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent-dry
observations from the flipper tag were preferred.

Field identification of age class can be inexact, particularly when discerning
subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges
[@mclaren1958a], and pelage characteristics for some species. In the case of
ribbon seals, subadults often have less distinct ribbons than adults. Bearded
seal subadults will often have a spotted pattern in the pelage that is not seen
in adults. Spotted seal pelage cannot be used to reliably discern adults from
subadults. Despite these challenges, we feel the age classifications used in
this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are likely sexually mature and older than
four years. For those bio-loggers deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the
following year. Table \@ref(tab:tblDeploy) provides a summary of these
deployments and data received from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land can still send data to satellites; i.e., a detached bio-logger 
that is dry (either on ice or land) will record and transmit data suggesting
the seal is hauled out. As such, end times of each deployment were identified by
examining bio-logger locations, percent-dry records, and dive behavior (if
available) to determine when bio-loggers ceased providing data consistent with
seal behavior. Data outside of the deployment start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible hourly percent-dry data. All data
after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
study season for each species. Observations for ribbon and spotted seals are
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data are
transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Bearded seal deployments are more
evenly spread with fewer observations in May and June. The majority of bearded
seal deployments start later in the summer and by May bio-loggers have either
fallen off or their batteries are depleted.

```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May."}

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10%
inclusive of 0% and 100% and additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, generally, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period; however, hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (i.e. tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
seem to greatly bias inference about haul-out probabilities [@conn2012b].

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May. Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. For any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location. Any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and seals' availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution of weighted locations with available haul-out behavior data for each
species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America.
Numerous weather variables are made available across the region 8 times daily.
For this study, NARR weather values were subset to the extent of our study area
over the Bering and Chukchi seas at 3-hr intervals based on the native grid
resolution of 32km (1024km^2^). The following meteorological variables are
known to affect haul-out behavior in other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal locations
using a bilinear method: 1) air temperature at 2m above the earth's surface, 2)
wind consisting of northerly and easterly vector components converted to wind
speed using the Euclidean norm, 3) barometric pressure at sea level, and 4)
precipitation (Table \@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models because, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

@@

We assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out behavior. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15% concentration for each year.

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation and for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
cubic effects of day-of-year to represent temporal changes in behavior. However,
unlike previous models for harbor seals [@london2012] and ice-associated seals
[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
capture diurnal cycles, we adopted a continuous formulation based on Fourier
series that provides a flexible model while preserving the inherent circularity
needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
also represents hour-of-day with 6 parameters, which is a considerable reduction
when compared to a 24 parameter variable, especially when fitting models with
interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification
for hour-of-day effects:

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages. Overall, bearded seals were less likely to haul out and had
a bi-modal distribution of haul-out probability across the day (Figure
\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
bearded seal model predicts additional haul-out activity around local midnight.
In concert with the lower magnitude of haul-out probability, bearded seal
haul-out behavior is also more protracted throughout the spring season compared
to ribbon and spotted seals.

>>>>

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering, Chukchi seas as a platform for important life history functions,
such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance of these species are unknown, so it is difficult to assess the effect,
if any, declines in sea ice habitat have had, or will have, on seal demography.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs during winter and spring.

The remoteness of the Bering and Chukchi seas means direct scientific observation
of seal behavior is impractical. Thus, bio-logging devices are especially useful
tools for collecting key information on movement and haul-out behavior for these
species. Bio-logging records of time spent out of the water provide valuable
data for identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about life history and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we address some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
accounted for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
variability in haul-out timing and possible linkage to changes in the extent of
seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability, and
investigates the potential impact of a changing icescape on the behavior of
these species.

@@

Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers were attached during studies based
in coastal Alaska communities or on research ships. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
to the primary literature for detailed capture and bio-logger attachment methods
(Supplemental Material, Table).

@@

Haul-out behavior data from 
`r species_n %>% .$n %>% sum()` bio-loggers deployed on 
`r species_n %>% filter(species == "Bearded seal") %>% .$n` bearded, 
`r species_n %>% filter(species == "Ribbon seal") %>% .$n` ribbon, and 
`r species_n %>% filter(species == "Spotted seal") %>% .$n` spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached through the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is floating at the surface, inducing
a slight positive bias in the hourly percent-dry values reported by the
bio-logger. For this study, in cases where both bio-logger types were deployed,
hourly percent-dry observations from the flipper tag were preferred.

Field identification of age class can be inexact, particularly when differentiating
subadults from adults. Sex as well as age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges
[@mclaren1958a], and pelage characteristics for some species. In the case of
ribbon seals, subadults often have less distinct ribbons than adults. Bearded
seal subadults will often have a spotted pattern in the pelage that is not seen
in adults. Spotted seal pelage cannot be used to reliably discern adults from
subadults. Despite these challenges, we feel the age classifications used in
this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are individuals that are likely sexually
mature and older than four years. For those bio-loggers deployed on
young-of-the-year and transmitting into the next year, the age class was
advanced to subadult on 1 March of the following year. Table
\@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt, attachment failure, or seal mortality and remain
on ice or land may continue to send data to satellites; i.e., a detached bio-logger 
that is dry (either on ice or land) will record and transmit data suggesting
the seal is hauled out. Therefore, end times of each deployment were identified by
examining bio-logger locations, percent-dry records, and dive behavior (if
available) to determine when bio-loggers ceased providing data consistent with
seal behavior. Data outside of the deployment start and end times were discarded
prior to analysis. After approximately 9 months, 7 devices deployed on the rear
flipper of bearded seals reported implausible hourly percent-dry data. All data
after the first instance of unrealistic values were censored from this analysis.
Figure \@ref(fig:dataCal) shows the distribution of all haul-out data across the
study season for each species. Observations for ribbon and spotted seals are
concentrated in the months of May and June due to the timing of deployment
(April and May) and the timing of molt (May and June). During molt, seals (and
their attached bio-loggers) spend more time out of the water and more data are
transmitted. Molt timing also impacts when many deployments end as any
bio-loggers adhered to the hair will fall off. Relative to the other species in
the study, there were fewer deployments of bio-loggers on bearded seals. This
resulted in fewer data observations overall and noticeably lower in numbers May
and June. The majority of bearded seal deployments start later in the summer and
by May bio-loggers have either fallen off or their batteries are depleted.

```{r dataCal, include=TRUE, fig.height = 4, fig.cap = "Distribution of hourly percent-dry bio-logger data from 1 March to 15 July for each species. Data are grouped by day-of-year and presented in seal-hours collated across all years between 2005 and 2020. The higher density of data from May and June in ribbon and spotted seals coincides with peak molting when seals (and their attached bio-loggers) are more likely hauled out. Additionally, many bio-logger deployments started in April and May. The overall reduced quantity of observations from bearded seals is reflective of the lower number of bio-logger deployments in the study."}

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10%
inclusive of 0% and 100% and additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, typically, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period, though hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
substantially bias inference about haul-out probabilities [@conn2012b].

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May. Areas of the plot indicated with an 'X' represent missing data that were not successfully recieved from the tag."}

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognized that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land are likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made
available across the region 8 times daily. For this study, NARR weather values
were subset to the extent of our study area over the Bering and Chukchi seas at
3-hr intervals based on the native grid resolution of 32km (1024km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
locations using a bilinear method: 1) air temperature at 2m above the earth's
surface, 2) wind consisting of northerly and easterly vector components
converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
level, and 4) precipitation (Table \@ref(tab:tblCovariates)).

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range during the spring and we were interested in possible differences in the
timing of haul-out behavior along a latitudinal gradient. We omitted the
latitudinal effect from ribbon and spotted seal models because, during the
spring, these species are most prevalent near the southern ice edge in the
Bering Sea [@conn2014a].

@@

We assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out probability. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15% concentration for each year.

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation and for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - round((portion_10(analysis_data) + portion_90(analysis_data))) * 100)`% of our
observations fell between 10% and 90% hourly percent-dry.

We followed Ver Hoef et al. [-@verhoef2014a] in using linear, quadratic, and
cubic effects of day-of-year to represent temporal changes in behavior. However,
unlike previous models for harbor seals [@london2012] and ice-associated seals
[@verhoef2014a], which treated hour-of-day as a 24-level categorical variable to
capture diurnal cycles, we adopted a continuous formulation based on Fourier
series that provides a flexible model while preserving the inherent circularity
needed for time-of-day effects (i.e., hour 0 should be equal to hour 24). It
also represents hour-of-day with 6 parameters, which is a considerable reduction
when compared to a 24-parameter variable, especially when fitting models including
interactions between hour-of-day and other variables (e.g., age-sex class,
day-of-year). According to this approach, we used the following specification
for hour-of-day effects:

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal)). Overall, bearded seals were
less likely to haul out and had a bi-modal distribution of haul-out probability
across the day. In addition to a peak around local solar noon, the bearded seal
model predicts additional haul-out activity around local midnight. In concert
with the lower magnitude of haul-out probability, bearded seal haul-out behavior
is also more protracted throughout the spring season compared to ribbon and
spotted seals.
